---
title: "Intro to data (lab 1 for BIOS 26122)"
author: "Dmitry Kondrashov"
format: html
editor: visual
---

## Description

The goal of this assignment is to introduce basic data description and visualization using base R and tidyverse packages. Specifically, you should be able to:

1.  Read in a data set as a data frame (or tibble) and report the number and types of variables, as well as the number of observations.
2.  Visualize a variable using a method appropriate for its type; observe the general shape of the distribution.
3.  Report descriptive statistics (e.g. counts, means, standard deviations) for numeric variables and generate cross tables for categorical variables.
4.  Perform an assessment of missing values and outliers and do basic data cleaning.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
