---
title: "Comparison of Hot and Cold Tumors"
author: "Gepoliano Chaves (Ph. D.)"
date: "December 16th, 2023"
#output: revealjs::revealjs_presentation
#output: html_notebook
#output: beamer_presentation
#output: github_document
#output: html_vignette
#output: ioslides_presentation
#output: slidy_presentation 
#output: tufte::tufte_book
output: pdf_document
---

To prepare material for this notebook I used notebooks ADRN-MES-Sengupta-Final.Rmd and "Survial MES.Rmd".

# 1) Library installation

```{r, some libraries, message=F, warning=F}
library(ggplot2)
library(GSVA)
library(GSVAdata)
library(data.table)
library(stringr)
library(tibble)
library(viridis)
library(expss)
library(ggpubr)
```

# 2) Gene Expression and GSVA

Load data-frames

```{r, load data-frame, warning=F, message=F}
GSE49711 <- readRDS("../../../../../Hypoxia_Paper/for loop MES/data/GSE49711.rds")
GSE49711_GeneMatrix <- as.matrix(GSE49711)
```

Load gene sets

```{r}
cfDNA_PCA_gene_list <- getGmt("~/Desktop/Gepoliano/Analysis/cfDNA_genes_PCA.txt")
```

Construct GSVA data-frame

```{r Construct GSVA data-frame}
# Needs to re-run this step if there are new gene sets
GSE49711_GSVA <- gsva(GSE49711_GeneMatrix, 
                          cfDNA_PCA_gene_list, 
                          min.sz=1, max.sz=Inf, 
                          verbose=TRUE)
```

Merge Gene Expression and GSVA data-frames

```{r, Merge Gene Expression and GSVA data-frames}
GSVA_GeneExpression_Matrix <- rbind(GSE49711_GSVA, GSE49711_GeneMatrix)
```

# 2) Process GSVA DF with Phenotype Scores

Previous chunk used was from notebook "Multiple Test Correction Kocak.Rmd". Below we will transpose and Name Column with Sample Names

```{r, 2) Process GSVA DF with Phenotype Scores}
GSVA_GeneExpression_Matrix_t <- t(GSVA_GeneExpression_Matrix)
GSVA_GeneExpression_Matrix_t_df <- as.data.frame(GSVA_GeneExpression_Matrix_t)
GSVA_GeneExpression_Matrix_t_df["Sample_ID"] <- rownames(GSVA_GeneExpression_Matrix_t_df)
```

## 2.1) Inflammatory Response vs Hot

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        HALLMARK_INFLAMMATORY_RESPONSE, 
        results_immunogenicity_hot,
        xlab = "HALLMARK_INFLAMMATORY_RESPONSE",
        ylab = "results_immunogenicity_hot",
      main = "Hallmark Inflammation vs. Hot")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.2) Inflammatory Response vs Cold

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        HALLMARK_INFLAMMATORY_RESPONSE, 
        results_immunogenicity_cold,
        xlab = "HALLMARK_INFLAMMATORY_RESPONSE",
        ylab = "results_immunogenicity_cold",
      main = "Hallmark Inflammation vs. Cold")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.3) Inflammatory Response vs ADRN

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        HALLMARK_INFLAMMATORY_RESPONSE, 
        List_RNA_Normoxia_ADRN_vs_MES_Up.txt,
        xlab = "HALLMARK_INFLAMMATORY_RESPONSE",
        ylab = "ADRN (1298)",
      main = "Hallmark Inflammation vs. ADRN (1298)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.4) Inflammatory Response vs MES

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        HALLMARK_INFLAMMATORY_RESPONSE, 
        List_RNA_Normoxia_ADRN_vs_MES_Down.txt,
        xlab = "HALLMARK_INFLAMMATORY_RESPONSE",
        ylab = "MES (1469)",
      main = "Hallmark Inflammation vs. MES (1469)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.5) Inflammatory Response vs ADRN_Norm_vs_Hypo_Up_554.txt

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        HALLMARK_INFLAMMATORY_RESPONSE, 
        ADRN_Norm_vs_Hypo_Up_554.txt,
        xlab = "HALLMARK_INFLAMMATORY_RESPONSE",
        ylab = "Hypoxia Up (554)",
      main = "Hallmark Inflammation vs. ADRN Hypoxia Up (554)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.6) Inflammatory Response vs ADRN_Norm_vs_Hypo_Down_635.txt

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        HALLMARK_INFLAMMATORY_RESPONSE, 
        ADRN_Norm_vs_Hypo_Down_635.txt,
        xlab = "HALLMARK_INFLAMMATORY_RESPONSE",
        ylab = "Hypoxia Down (635)",
      main = "Hallmark Inflammation vs. ADRN Hypoxia Down (635)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.7) Check MES vs. Hypoxia Up

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        List_RNA_Normoxia_ADRN_vs_MES_Down.txt, 
        ADRN_Norm_vs_Hypo_Up_554.txt,
        xlab = "MES (1469)",
        ylab = "Hypoxia Up (554)",
      main = "MES (1469) vs. ADRN Hypoxia Up (554)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.8) Check ADRN vs. Hypoxia Down

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        List_RNA_Normoxia_ADRN_vs_MES_Down.txt, 
        ADRN_Norm_vs_Hypo_Down_635.txt,
        xlab = "ADRN (1298)",
        ylab = "Hypoxia Down (635)",
      main = "ADRN (1298) vs. ADRN Hypoxia Down (635)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.9) Check Hot vs. Hypoxia Up

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        results_immunogenicity_hot, 
        ADRN_Norm_vs_Hypo_Up_554.txt,
        xlab = "results_immunogenicity_hot",
        ylab = "Hypoxia Up (554)",
      main = "Immunologicaly Hot vs. ADRN Hypoxia Up (554)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

## 2.10) Check Cold vs. Hypoxia Down

```{r}
qplot(data = GSVA_GeneExpression_Matrix_t_df,
        results_immunogenicity_cold, 
        ADRN_Norm_vs_Hypo_Down_635.txt,
        xlab = "results_immunogenicity_cold",
        ylab = "Hypoxia Down (635)",
      main = "Immunologicaly Cold vs. ADRN Hypoxia Down (635)")+
  geom_smooth(method=lm)+
  stat_cor(label.x = (-.3), label.y = 0.5) +
  stat_regline_equation(label.x = (-.3), label.y = 0.4)+
  theme_bw()
```

# Correlation Matrix

```{r, eval=F}
r2_gse62564_gsva_genes_df <- t(r2_gse62564_GSVA_genes)
```

# Select terms to be plotted

```{r}
GSVA_GeneExpression_Matrix <- GSVA_GeneExpression_Matrix_t_df[,
                                                              c("HALLMARK_INFLAMMATORY_RESPONSE",
                                                                "results_immunogenicity_hot",
                                                                "results_immunogenicity_cold",
                                                                "List_RNA_Normoxia_ADRN_vs_MES_Up.txt",
                                                                "List_RNA_Normoxia_ADRN_vs_MES_Down.txt",
                                                                "ADRN_Norm_vs_Hypo_Up_554.txt",
                                                                "ADRN_Norm_vs_Hypo_Down_635.txt"#, This part needs to be commented w/ part below
                                                                # "MES_Norm_vs_Hypo.compCas_Up_287.txt",
                                                                # "MES_Norm_vs_Hypo.compCas_Down_171.txt",
                                                                # "MES_Norm_vs_Hypo.compCas_Up_212.txt",
                                                                # "MES_Norm_vs_Hypo.compCas_Down_103.txt"
                                                        )]


GSVA_GeneExpression_Matrix_df <- as.data.frame(GSVA_GeneExpression_Matrix)

GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Inflam_Response = HALLMARK_INFLAMMATORY_RESPONSE)
GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Hot = results_immunogenicity_hot)
GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Cold = results_immunogenicity_cold)
GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(ADRN_1298 = List_RNA_Normoxia_ADRN_vs_MES_Up.txt)
GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(MES_1469 = List_RNA_Normoxia_ADRN_vs_MES_Down.txt)
GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Hypo_Up_554 = ADRN_Norm_vs_Hypo_Up_554.txt)
GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Hypo_Down_635 = ADRN_Norm_vs_Hypo_Down_635.txt)

# GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Hypo_UP_287 = MES_Norm_vs_Hypo.compCas_Up_287.txt)
# GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Hypo_DN_171 = MES_Norm_vs_Hypo.compCas_Down_171.txt)
# GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Hypo_UP_212 = MES_Norm_vs_Hypo.compCas_Up_212.txt)
# GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% dplyr::rename(Hypo_DN_103 = MES_Norm_vs_Hypo.compCas_Down_103.txt)
```

Make character, numeric

```{r}
library(dplyr)
GSVA_GeneExpression_Matrix_df <- GSVA_GeneExpression_Matrix_df %>% mutate_if(is.character, as.numeric)
```

# Calculate correlation

```{r}
GSVA_GeneExpression_Matrix_cor_df <- cor(GSVA_GeneExpression_Matrix_df)
```

# Plot correlation matrix

Plot with colorful number

```{r, corrplot colorful numbers}
library(corrplot )
corrplot(GSVA_GeneExpression_Matrix_cor_df, method = 'number', order="hclust", addrect = 3) 
```

Plot correlations as circles. This figure intentionally compares the correlations of cellular MES signatures and hot/cold tumors.

```{r corrplot circles}
corrplot(GSVA_GeneExpression_Matrix_cor_df, 
         type="lower",
         order="hclust",
         #method = c(pie),
         title = "Correlation Matrix Hot and Cold Tumors",
         tl.col = "black",
         #tl.offset=1,
         tl.srt = 45,
         addrect = 4,
         col=colorRampPalette(c("#053061",
                                "#2166AC",
                                "#4393C3",
                                "#92C5DE",
                                "#D1E5F0",
                                "#FFFFFF",
                                "#FDDBC7",
                                "#F4A582",
                                "#D6604D", 
                                "#B2182B",
                                "#67001F"
                                ))(200)
         )
```
